---
title: "Finding  modules and hub genes associated with S100A trait"
author: "Halfdan Rydbeck, Veronika Golubinskaya"
date: "3/11/2020"
site: bookdown::bookdown_site
#output: bookdown::gitbook#,output_dir:../book
output:
  bookdown::pdf_book:
    keep_tex: yes
    latex_engine: xelatex
  bookdown::gitbook:
    lib_dir: assets
    config:
      toolbar:
        position: static
  bookdown::tufte_html_book:
    split_by: chapter
    toc: yes
    css: toc.css
---
documentclass: book
bibliography: [./bib/web.bib,./bib/packages.bib,./bib/library_for_alarmins.bib]
biblio-style: apalike
link-citations: yes
keep_tex: yes
github-repo: rstudio/bookdown-demo
description: "This project is looking to characterize S100A genes or alarmins in terms of their involvement in co-expression modules and in terms of being hub genes/having high connectivity. It also want to  establish to what degree alarmins correlate with clinical outcomes"
---

#include = FALSE prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
#echo = FALSE prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
#message = FALSE prevents messages that are generated by code from appearing in the finished file.

# Introduction

## The R package WGCNA

### Website and publication
The R package WGCNA has its [website](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/) with [tutorials](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html), on which this pipeline is based, and also a page with [explanations of fundamental concepts](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/Simulated-00-Background.pdf). All steps in this analysis, but the last in which hub genes are filtered out, is based on Tutorial I. The step in which hubgenes are filtered out is based on [Tutorial III, section 7](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/Simulated-07-Membership.pdf). The publication of the software, see @RN1, has been cited  5454 times according to Google Scholar.

### What it does
WGCNA derives various measures of gene to gene relatedness based on expression profiles. The measures are used for clustering genes, creating gene co-expression modules and weighted gene-gene interaction networks. WGCNA also clusters samples (and their clinical outcomes) based on on their gene expression profiles. Further WGCNA identifies modules that are significantly correlated with clinical outcomes. Finally the R package identifies so called "hub genes" based on the number of genes that they have strong connections to and on the degree to which they correlate to a clinical outcome.

## Input 1 Gene expression
In this analysis we are looking at a table of gene expression values. In the table the rows represents children and the columns represents genes. Each row represents a gene expression profile of a child and each column represents a gene expression profile of a gene. There are 10533 genes and 33 children. **display some table here**. The expression profiles can be used to derive measures of relative distances or closeness between children and between genes. WGCNA uses these distances to perform hierarchical clustering of the genes and the children. It further uses the clustering to derive modules of genes with highly correlated expression profiles. Finally it also uses the distances between genes to derive weighted networks. The two  basic terminologies of networks are nodes and edges. In gene expression networks this translates to genes and the connections.

## Input 2 Clinical outcome

```{r eval=FALSE, include=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

```{r setup, include=FALSE}
# Searched for "tidyverse pdf r markdown" (https://chrisbeeley.net/?p=1037)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), './bib/packages.bib')
```
Loading libraries
```{r include=FALSE}
library("tidyverse")
#library("readxl")
#install.packages(c("naniar"))
#library(naniar)
```





